---
title: Effects of Dredging Related Sedimentation on Larval Settlement in Harbor Porites
  Coral CODE
author: "Sami"
date: "2024-05-01"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(MASS)
library(effects)
library(lmtest)
library(statmod)
library(plyr)

#Modify location as needed
data <- read.csv("~/Desktop/sediment_settlement_experiment/data/data.csv", header = TRUE)

```

# Data Cleaning

## Defining Data Frame
```{r, echo=FALSE, warning=FALSE}
#Look at data
head(data)

#Change trial to a factor from an integer
data$Trial <- as.factor(data$Trial)
head(data)

#Add a vector/column for natural log transformed concentration. Add a small value because of the presences of zeroes (controls) in the dataset.
data$logConc <- log(data$Treatment+0.0001)
head(data)

#Add column for total larvae
data$total_larvae <- data$settle_n + data$meta_n + data$swim_n +data$dead_n
head(data)

```

# Response: Settlement
  
## GLM Model Validation
```{r, echo=FALSE}
# Start with full, binomial model
mod1 <- glm(settle_prop ~ Treatment + Trial, family = binomial, weights = total_larvae,
            data = data)
summary(mod1)

# Extract residual deviance
residual_deviance <- deviance(mod1)

# Degrees of freedom
df <- nrow(model.matrix(mod1)) - length(coefficients(mod1))

# Compute p-value to test for overdispersion
p_value <- 1 - pchisq(residual_deviance, df)

# Print results
print(paste("Residual Deviance:", residual_deviance))
print(paste("Degrees of freedom:", df))
print(paste("P-value:", p_value))

```

Since there is evidence of overdispersion (p<0.05 in deviance goodness-of-fit test), let's make this a quasibinomial model and compare the fit/performance of several versions: full or reduced, and with or without log transformed concentration.

```{r, echo=FALSE, warning=FALSE}
mod2 <- glm(settle_prop ~ Treatment + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod3 <- glm(settle_prop ~ logConc + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod4 <- glm(settle_prop ~ Treatment, family = quasibinomial, weights = total_larvae,
            data = data)
mod5 <- glm(settle_prop ~ logConc, family = quasibinomial, weights = total_larvae,
            data = data)

# Compare full vs reduced models, first between non-log transformed and then between log transformed
anova(mod4, mod2, test = "Chisq")
anova(mod5, mod3, test = "Chisq") 

```

Now let's compare the model residuals for mod3 and mod4, the two preferred models in the Chi-sq (anova) tests above.

```{r, echo=FALSE, warning=FALSE}
#Residual plots for mod3
par(mfrow=c(2,2))
scatter.smooth(predict(mod3, type='response'), rstandard(mod3, type='deviance'), col='gray')
scatter.smooth(predict(mod3), resid(mod3, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod3, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod3, type='deviance'), col='gray')

#Residual plots for mod4
scatter.smooth(predict(mod4, type='response'), rstandard(mod4, type='deviance'), col='gray')
scatter.smooth(predict(mod4), resid(mod4, type='working'), col='gray')
scatter.smooth(data$Treatment, rstandard(mod4, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod4, type='deviance'), col='gray')

```

The residual plots look much cleaner for mod3 (~ logConc + Trial) than for mod4 (~ Conc), and it makes sense to look at the response as a function of log-transformed concentration given the concentration values: 0, 0.5, 1, 10, 25, and 50.

Now let's try to plot the model predictions.

## GLM Prediction Plots
```{r, echo=FALSE, warning=FALSE}
# Model predictions
mod3_eff <- effect("logConc", mod3)
mod3_eff <- as.data.frame(mod3_eff)

# Custom ticks and labels 
custom_ticks <- c(0.000125, 0.5, 1, 10, 25, 50)  
custom_labels <- c("0", "0.5", "1", "10", "25", "50")

# Plot using ggplot2
settle_plot <- ggplot(mod3_eff, aes(x = logConc, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(x = bquote(Sediment~Concentration~(mg/cm^2)), y = "Settlement Rate (%)") +
  scale_x_continuous(
    breaks = log(custom_ticks),
    labels = custom_labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1),
    labels = function(x) paste0(x*100)
  ) +
  theme_minimal()

#Model summary
summary(mod3)

```

## Interpreting Results

There is a significant decrease in larval settlement as sediment concentration increases.  Since sediment concentrations were log-transformed in the model, let's back-transform the model output to quantify the effect of sediment on settlement.  

If logit(pi) = beta0 + beta1*log(X), then a k-fold change in X corresponds to a multiplicative change in the odds of k^beta-1. Given this general formula, below are two ways to interpret the results:

1. for every one unit increase in sediment concentration
2. for every ten unit increase in sediment concentration

Let's start with a one unit increase in concentration...

```{r, echo=FALSE, warning=FALSE}
#If logit(pi) = beta0 + beta1*log(X), then a k-fold change in X corresponds to a multiplicative change in the odds of k^beta-1.

# Extract 95% confidence intervals for the coefficients
conf_intervals <- confint(mod3, level = 0.95)

# Extract coefficient estimates
coefficients <- coef(mod3)

# Back-transform the coefficients and confidence intervals
exp_coefficients <- exp(coefficients) - 0.0001 
exp_conf_intervals <- exp(conf_intervals) - 0.0001

# Extract p-values from the model summary
p_values <- round(summary(mod3)$coefficients[, 4], digits = 3)  # Column 4 contains p-values

# Create a data frame with back-transformed estimates, confidence intervals, and p-values
estimates <- cbind(exp_coefficients, exp_conf_intervals, p_values)

# Print the estimates
print(estimates)

# Because these estimates are less than one, let's subtract from 1 to get the % decrease.
dec_exp_coefficients <- round(1 - (exp(coefficients) - 0.0001), digits = 3)
dec_exp_conf_intervals <- round(1 - (exp(conf_intervals) - 0.0001), digits = 3)

# Create another data frame with estimates of percent/proportion decline, associated confidence intervals, and p-values
estimates2 <- cbind(dec_exp_coefficients, dec_exp_conf_intervals, p_values)

# Print the estimates
print(estimates2)

```

The estimated odds of a larva settling decrease by 16.7% for every one unit increase in sediment concentration (95% CI 11.5-21.8, GLM p<0.001).

Now, let's find out what the change is for every ten unit increase in concentration...
  
```{r, echo=FALSE, warning=FALSE}
#If logit(pi) = beta0 + beta1*log(X), then a k-fold change in X corresponds to a multiplicative change in the odds of k^beta-1.

# Back-transform the coefficients and confidence intervals
ten_coefficients <- round(exp(10*coefficients) - 0.0001, digits = 3)
ten_conf_intervals <- round(exp(10*conf_intervals) - 0.0001, digits = 3)

# Create a data frame with back-transformed estimates, confidence intervals, and p-values
ten_estimates <- cbind(ten_coefficients, ten_conf_intervals, p_values)

# Print the estimates
print(ten_estimates)

# Because these estimates are less than one, let's subtract from 1 to get the % decrease.
dec_ten_coefficients <- round(1 - (exp(10*coefficients) - 0.0001), digits = 3)
dec_ten_conf_intervals <- round(1 - (exp(10*conf_intervals) - 0.0001), digits = 3)

# Create another data frame with estimates of percent/proportion decline, associated confidence intervals, and p-values
ten_estimates2 <- cbind(dec_ten_coefficients, dec_ten_conf_intervals, p_values)

# Print the estimates
print(ten_estimates2)

```

For every ten unit increase in sediment concentration, the estimated odds of a larva settling decrease by 83.8% (95% CI 70.4-91.4, GLM p<0.001).

# Response: Partially metamorphosed
  
## GLM Model Validation

```{r, echo=FALSE}
# Start with full, binomial model
mod1meta <- glm(meta_prop ~ Treatment + Trial, family = binomial, weights = total_larvae,
            data = data)
summary(mod1meta)

# Extract residual deviance
residual_deviance_meta <- deviance(mod1meta)

# Degrees of freedom
df_meta <- nrow(model.matrix(mod1meta)) - length(coefficients(mod1meta))

# Compute p-value to test for overdispersion
p_value_meta <- 1 - pchisq(residual_deviance_meta, df)

# Print results
print(paste("Residual Deviance:", residual_deviance_meta))
print(paste("Degrees of freedom:", df_meta))
print(paste("P-value:", p_value_meta)) 

```

There is no evidence of overdispersion with the binomial, so I will keep the binomial.
  
```{r, echo=FALSE, warning=FALSE}
mod2meta <- glm(meta_prop ~ Treatment + Trial, family = binomial, weights = total_larvae,
            data = data)
mod3meta <- glm(meta_prop ~ logConc + Trial, family = binomial, weights = total_larvae,
            data = data)
mod4meta <- glm(meta_prop ~ Treatment, family = binomial, weights = total_larvae,
            data = data)
mod5meta <- glm(meta_prop ~ logConc, family = binomial, weights = total_larvae,
            data = data)

# Compare full vs reduced models, first between non-log transformed and then between log transformed
anova(mod4meta, mod2meta, test = "Chisq") 
anova(mod5meta, mod3meta, test = "Chisq") 

```

Based on the Chi sq test, mod2 and mod3 are preferred.  Let's plot the residuals.  

```{r, echo=FALSE, warning=FALSE}
#Residual plots for mod2
par(mfrow=c(2,2))
scatter.smooth(predict(mod2meta, type='response'), rstandard(mod2meta, type='deviance'), col='gray')
scatter.smooth(predict(mod2meta), resid(mod2meta, type='working'), col='gray')
scatter.smooth(data$Treatment, rstandard(mod2meta, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod2meta, type='deviance'), col='gray')

#Residual plots for mod3
scatter.smooth(predict(mod3meta, type='response'), rstandard(mod3meta, type='deviance'), col='gray')
scatter.smooth(predict(mod3meta), resid(mod3meta, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod3meta, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod3meta, type='deviance'), col='gray')

```

The residual plots look cleaner for mod3 (~ logConc + Trial) than for mod2 (~ Conc + Trial), so let's look at the response as a function of log-transformed concentration.  

## GLM Prediction Plots

```{r, echo=FALSE, warning=FALSE}
# Model predictions
mod3meta_eff <- effect("logConc", mod3meta)
mod3meta_eff <- as.data.frame(mod3meta_eff)

# Custom ticks and labels
custom_ticks <- c(0.000125, 0.5, 1, 10, 25, 50)
custom_labels <- c("0", "0.5", "1", "10", "25", "50")

# Plot using ggplot2 with natural log-scaled x-axis, custom ticks, and labels
pm_plot <- ggplot(mod3meta_eff, aes(x = logConc, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(x = bquote(Sediment~Concentration~(mg/cm^2)), y = "Proportion partially metamorphosed") +
  scale_x_continuous(
    breaks = log(custom_ticks),
    labels = custom_labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1)
  ) +
  theme_minimal()

```

# Interpreting results 
The model has wide confidence intervals and the predicted values are consistently different from the actual values. Instead, let's visualize the raw data for partial metamorphosis.

```{r}
meta_plot <- ggplot(data, aes(x = logConc, y = meta_prop)) +
  geom_point(position = "jitter") +
  geom_smooth(method = "lm", se = FALSE, color = "black") + 
  labs(x = bquote(Sediment~Concentration~(mg/cm^2)), y = "Partial Metamorphosis Rate (%)") +
  scale_x_continuous(
    breaks = log(custom_ticks),
    labels = custom_labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1),
    labels = function(x) paste0(x*100)
  ) +
  theme_minimal()

```

# Response: Swimming
  
## GLM Model Validation

```{r, echo=FALSE}
# Start with full, binomial model
mod1swim <- glm(swim_prop ~ Treatment + Trial, family = binomial, weights = total_larvae,
            data = data)
summary(mod1swim)

# Extract residual deviance
residual_deviance_swim <- deviance(mod1swim)

# Degrees of freedom
df_swim <- nrow(model.matrix(mod1swim)) - length(coefficients(mod1swim))

# Compute p-value to test for overdispersion
p_value_swim <- 1 - pchisq(residual_deviance_swim, df)

# Print results
print(paste("Residual Deviance:", residual_deviance_swim))
print(paste("Degrees of freedom:", df_swim))
print(paste("P-value:", p_value_swim)) 

```

There is evidence of overdispersion (p < 0.05) so let's change this to a quasibinomial distribution.  

```{r, echo=FALSE, warning=FALSE}
mod2swim <- glm(swim_prop ~ Treatment + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod3swim <- glm(swim_prop ~ logConc + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod4swim <- glm(swim_prop ~ Treatment, family = quasibinomial, weights = total_larvae,
            data = data)
mod5swim <- glm(swim_prop ~ logConc, family = quasibinomial, weights = total_larvae,
            data = data)

# Compare full vs reduced models, first between non-log transformed and then between log transformed
anova(mod4swim, mod2swim, test = "Chisq") 
anova(mod5swim, mod3swim, test = "Chisq") 

```

Based on the Chi sq test, mod2 and mod3 are preferred.  Let's plot the residuals.  

```{r, echo=FALSE, warning=FALSE}
#Residual plots for mod2
par(mfrow=c(2,2))
scatter.smooth(predict(mod2swim, type='response'), rstandard(mod2swim, type='deviance'), col='gray')
scatter.smooth(predict(mod2swim), resid(mod2swim, type='working'), col='gray')
scatter.smooth(data$Treatment, rstandard(mod2swim, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod2swim, type='deviance'), col='gray')

#Residual plots for mod3
scatter.smooth(predict(mod3swim, type='response'), rstandard(mod3swim, type='deviance'), col='gray')
scatter.smooth(predict(mod3swim), resid(mod3swim, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod3swim, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod3swim, type='deviance'), col='gray')

```

The residual plots look cleaner for mod3 (~ logConc + Trial) than for mod2 (~ Conc + Trial), so let's look at the response as a function of log-transformed concentration.  

## GLM Prediction Plots

```{r, echo=FALSE, warning=FALSE}
# Model predictions
mod3swim_eff <- effect("logConc", mod3swim)
mod3swim_eff <- as.data.frame(mod3swim_eff)

# Custom ticks and labels
custom_ticks <- c(0.000125, 0.5, 1, 10, 25, 50) 
custom_labels <- c("0", "0.5", "1", "10", "25", "50")

# Plot using ggplot2 with natural log-scaled x-axis, custom ticks, and labels
swim_plot <- ggplot(mod3swim_eff, aes(x = logConc, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(x = bquote(Sediment~Concentration~(mg/cm^2)), y = "Rate of Larval Swimming (%)") +
  scale_x_continuous(
    breaks = log(custom_ticks),
    labels = custom_labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1),
    labels = function(x) paste0(x*100)
  ) +
  theme_minimal()

```

# Interpreting results 
The estimated odds of a larva swimming was unaffected by sediment concentration (GLM p > 0.05).

# Response: Mortality
  
## GLM Model Validation

```{r, echo=FALSE}
# Start with full, binomial model (log-transformed?)
mod1dead <- glm(dead_prop ~ Treatment + Trial, family = binomial, weights = total_larvae,
            data = data)
summary(mod1dead)

# Extract residual deviance
residual_deviance_dead <- deviance(mod1dead)

# Degrees of freedom
df_dead <- nrow(model.matrix(mod1dead)) - length(coefficients(mod1dead))

# Compute p-value to test for overdispersion
p_value_dead <- 1 - pchisq(residual_deviance_dead, df)

# Print results
print(paste("Residual Deviance:", residual_deviance_dead))
print(paste("Degrees of freedom:", df_dead))
print(paste("P-value:", p_value_dead))  

```

There is evidence of overdispersion (p < 0.05) so let's change this to a quasibinomial distribution.  

```{r, echo=FALSE, warning=FALSE}
mod2dead <- glm(dead_prop ~ Treatment + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod3dead <- glm(dead_prop ~ logConc + Trial, family = quasibinomial, weights = total_larvae,
            data = data)
mod4dead <- glm(dead_prop ~ Treatment, family = quasibinomial, weights = total_larvae,
            data = data)
mod5dead <- glm(dead_prop ~ logConc, family = quasibinomial, weights = total_larvae,
            data = data)

# Compare full vs reduced models, first between non-log transformed and then between log transformed
anova(mod4dead, mod2dead, test = "Chisq") 
anova(mod5dead, mod3dead, test = "Chisq") 

```

 Based on the Chi sq test, mod4 and mod5 are sufficient, but since this deviates from the models for the previous larval responses, let's take a look at the residuals for mod2, mod3, mod4 and mod5.  

```{r, echo=FALSE, warning=FALSE}
#Residual plots for mod4
par(mfrow=c(2,2))
scatter.smooth(predict(mod4dead, type='response'), rstandard(mod4dead, type='deviance'), col='gray')
scatter.smooth(predict(mod4dead), resid(mod4dead, type='working'), col='gray')
scatter.smooth(data$Treatment, rstandard(mod4dead, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod4dead, type='deviance'), col='gray') #note Trial not in mod5

#Residual plots for mod5
scatter.smooth(predict(mod5dead, type='response'), rstandard(mod5dead, type='deviance'), col='gray')
scatter.smooth(predict(mod5dead), resid(mod5dead, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod5dead, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod5dead, type='deviance'), col='gray') #note: Trial not in mod5

#Residual plots for mod3
scatter.smooth(predict(mod3dead, type='response'), rstandard(mod3dead, type='deviance'), col='gray')
scatter.smooth(predict(mod3dead), resid(mod3dead, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod3dead, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod3dead, type='deviance'), col='gray') 

#Residual plots for mod2
scatter.smooth(predict(mod2dead, type='response'), rstandard(mod2dead, type='deviance'), col='gray')
scatter.smooth(predict(mod2dead), resid(mod2dead, type='working'), col='gray')
scatter.smooth(data$logConc, rstandard(mod2dead, type='deviance'), col='gray')
scatter.smooth(data$Trial, rstandard(mod2dead, type='deviance'), col='gray') #note: Trial not in mod2

```

The residual plots for mod3 look better, and to keep models consistent across response types, let's look at the response as a function of log-transformed concentration.  


```{r, echo=FALSE, warning=FALSE}
# Model predictions
mod3dead_eff <- effect("logConc", mod3dead)
mod3dead_eff <- as.data.frame(mod3dead_eff)

# Custom ticks and labels
custom_ticks <- c(0.000125, 0.5, 1, 10, 25, 50)  
custom_labels <- c("0", "0.5", "1", "10", "25", "50")

# Plot using ggplot2 with natural log-scaled x-axis, custom ticks, and labels
dead_plot <- ggplot(mod3dead_eff, aes(x = logConc, y = fit)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) +
  labs(x = bquote(Sediment~Concentration~(mg/cm^2)), y = "Larval Mortality Rate (%)") +
  scale_x_continuous(
    breaks = log(custom_ticks),
    labels = custom_labels
  ) +
  scale_y_continuous(
    breaks = seq(0, 1, 0.2),
    limits = c(0, 1),
    labels = function(x) paste0(x*100)
  ) +
  theme_minimal()

```

# Interpreting results 
The estimated odds of a larva dying was unaffected by sediment concentration (GLM p > 0.05).

